\large{\bf Abnormal quantitative genetics? }

### 2023-07-03 writeup.tex
.

%%%%%%%%%%%%%%%%%%%%%%
\section{Non-Gaussian effects}

@@

\begin{proof}[of Lemma~\ref{lem:nongaussian_nonindep}]
    \comment{calculations with characteristic functions?}
\end{proof}

This result should perhaps not be surprising:
there is a deep connection between independence and the Gaussian distribution.
For instance, a proof of the basic properties of the (Gaussian) infinitesimal model
might be sketched as follows:
divide the parental loci into groups according to the results of inheritance,
defining
\begin{align*}
    U_M^{[i]} = \sum_{\ell=1}^M X_\ell \eta_\ell^{[i]} 
    \qquad \text{ and } \qquad
    V_M^{[i]} = \sum_{\ell=1}^M (1-X_\ell) \eta_\ell^{[i]} ,
\end{align*}
so that 
$Z_M^{[1]} = U_M^{[1]} + V_M^{[1]}$ and
$Z_M^{[2]} = U_M^{[2]} + V_M^{[2]}$, so
\begin{align*}
    Z_M = U_M^{[1]} + V_M^{[2]} \qquad \text{ and } \qquad
    R_M = \frac{U_M^{[1]} - V_M^{[1]} - U_M^{[2]} + V_M^{[2]}}{ 2 } .
\end{align*}
Suppose we then can prove (as is indeed true in the Gaussian case) that
(a) $(U_M^{[1]}, V_M^{[1]}, U_M^{[2]}, V_M^{[2]})$ converge in distribution;
(b) the limiting random variables are independent; and
defining $R$, $Z^{[1]}$ and $Z^{[2]}$ as is natural in the limit,
(c) $R$ is independent of $Z^{[1]}$ and $Z^{[2]}$.
The Kac-Bernstein theorem states that if two independent random variables
have the property that if their sum and difference are independent of each other,
then the random variables have a Gaussian distribution
(\citet{kac1939characterization,bernstein1941property},
but see \citet{kagan1973characterization} for many much more general results).
It is not hard to see that (c) implies that $U^{[1]}$ and $V^{[1]}$
are such a pair of random variables, and therefore Gaussian.

Is there a converse? In other words,
does independence of segregation noise imply a Gaussian distribution of trait values?
It is not clear to us in how much generality such a statement might be made,
but we can at least prove the following:

\begin{lemma}\label{lem:parental_contribs}
    Under the conditions of Lemma~\ref{lem:nongaussian_nonindep},
    $U^{[1]}$ and $V^{[1]}$ are not independent.
    \comment{OR SOMETHING?}
\end{lemma}

\begin{proof}[of Lemma~\ref{lem:parental_contribs}]
    \comment{more calcs with characteristic functions?}
\end{proof}

>>>>

%%%%%%%%%%%%%%%%%%%%%%
\section{Independence and the Gaussian assumption}

@@

\begin{proof}[Proof of Lemma~\ref{lem:nongaussian_nonindep}]
    \comment{calculations with characteristic functions?}
\end{proof}

Is there a converse? In other words,
does independence of segregation noise imply a Gaussian distribution of trait values?
It is not clear to us in how much generality such a statement might be made,
but we can at least prove the following.
Divide the parental loci into groups according to the results of inheritance,
defining
\begin{align*}
    U_M^{[i]} = \sum_{\ell=1}^M X_\ell \eta_\ell^{[i]} 
    \qquad \text{ and } \qquad
    V_M^{[i]} = \sum_{\ell=1}^M (1-X_\ell) \eta_\ell^{[i]} ,
\end{align*}
so that 
$Z_M^{[1]} = U_M^{[1]} + V_M^{[1]}$ and
$Z_M^{[2]} = U_M^{[2]} + V_M^{[2]}$, so
\begin{align*}
    Z_M = U_M^{[1]} + V_M^{[2]} \qquad \text{ and } \qquad
    R_M = \frac{U_M^{[1]} - V_M^{[1]} - U_M^{[2]} + V_M^{[2]}}{ 2 } .
\end{align*}
It turns out that independence of the portions inherited and not inherited by the offspring
implies a Gaussian distribution:

\begin{lem}\label{lem:parental_contribs}
    Suppose that
    $(U_M^{[1]}, V_M^{[1]}, U_M^{[2]}, V_M^{[2]})$ converge in distribution as $M \to \infty$
    to independent random variables $(U^{[1]}, V^{[1]}, U^{[2]}, V^{[2]})$
    (and so $Z_M$, $R_M$, $Z_M^{[1]}$ and $Z_M^{[2]}$ converge to $Z$, $R$, $Z^{[1]}$ and $Z^{[2]}$
    defined in the natural way).
    Then if $R$ is independent of $Z^{[1]}$ and $Z^{[2]}$,
    $(U^{[1]}, V^{[1]}, U^{[2]}, V^{[2]})$ are jointly Gaussian.
\end{lem}

\begin{proof}[Proof of Lemma~\ref{lem:parental_contribs}]
    The Kac-Bernstein theorem states that if two independent random variables
    have the property that if their sum and difference are independent of each other,
    then the random variables have a Gaussian distribution
    (\citet{kac1939characterization,bernstein1941property},
    but see \citet{kagan1973characterization} for many much more general results).
    The assumptions imply that $U^{[1]}$ and $V^{[1]}$
    are such a pair of random variables, and therefore Gaussian.
\end{proof}

It is clear in the Gaussian case
that $U^{[1]}$ and $V^{[1]}$ are asymptotically independent.
More generally, independence depends on how the model is set up.
If the allelic effects $\eta_\ell^{[i]}$ are i.i.d.,
then even for finite $M$, the parental contributions
$(U_M^{[1]}, V_M^{[1]}, U_M^{[2]}, V_M^{[2]})$ are independent.
If the distribution of $\eta_\ell^{[i]}$ is chosen to obtain a non-Gaussian limit,
then Lemma~\ref{lem:parental_contribs} implies that $R$ cannot be independent of the parental traits.

However, the presence of a large effect polymorphic allele
makes $(U^{[1]}, V^{[1]}, U^{[2]}, V^{[2]})$ dependent:
consider the extreme case where $\eta^{[i]}_\ell$ are i.i.d.{} with variance 1 for $\ell \ge 2$,
while $\eta^{[i]}_1 = \pm 10$ with probability 1/2 each,
and $Z^{[1]} = \eta_1^{[1]} + \sum_{\ell=2}^M \eta_\ell^{[1]} / \sqrt{M}$.
Then in the limit $U^{[i]}$ and $V^{[i]}$ are Gaussian with mean either 0 or $\pm 10$ and variance 1/2,
and are highly dependent: $|U^{[i]}| \approx 10$ or $|V^{[i]}| \approx 10$ but not both.
(Segregation noise, meanwhile, depends strongly on parental trait values:
if $Z^{[1]} \approx Z^{[2]}$ then most likely
$\eta_1^{[1]} = \eta_1^{[2]}$ and so $R$ is Normal with variance 1/2;
but in the other case then $R$ has equal probabilities of being close to $+5$ or $-5$.)
